---
title: "Exercise lesson 2"
author: "Blanca Lizarbe"
date: "20/10/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
library(knitr)
```

# **Exercise Chapter 2**: Performing a manual normal q-q plot

We are going to use the data from Chapter 2 (P53) to compare a manual qqplot with R-function based qqplots. Remember that qqplots allow the visualization of the real experimental data distribution (y axis) vs. the quantiles of the normal standard function (x axis).

Create the following qqplots :

1. Using the plot function, plot in the y axis the (0,25,50,75 and 100%) quantiles of the experimental data p53 (you can use function "quantile") Vs the same quantiles of the standard normal distribution
2. Using the plot function, plot in the y axis the experimental data (sorted) and in the x axis the manual calculation of the quantiles (for the equivalent experimental "n" points of the p53 data)of a standard normal distribution. Add a 45ยบ line. 
3. a plot using the R function "**qqnorm**". Using the **qqline**, add the 45ยบ line. 
4. a plot using **qqPlot**

## 1. Plotting the (0,25%,50%,75% and 100%) quantiles 
```{r, echo = TRUE}

dp53 <- read.table("C:/Users/Laboratorio/Dropbox/PC (2)/Documents/UAM 23-24/BM-2/Lesson2/Lesson-2-files/P53.txt", header = TRUE, stringsAsFactors = TRUE)

# the function quantiles(x) provides the inverse value of the probability (or 
# "area under the curve"). It is the "cutoff" value at which a specific 
# percentage of the data lays on. By default, it provides the quantiles every 
# 0.25 step, from 0 to 1.

y <- quantile(dp53$p53)
y
# By observing the obtained "y", we can deduce for example that 25% of the data 
# of p53 have lower values than 1.9395

# for the standard normal distribution, we can use directly the qnorm function.
# There is no need (as many of you did) of generating a specific dataset normaly
# distributed around 0. The qnorm function assumes that, and by providing the 
# specific probabilities (0, 0.25...) it gives you the quantiles.
x <- qnorm(c(0.01,0.25,0.5,0.75,0.99))
# I set the first and last probabilities as 0.01 & 0.99 because 0 and 1 lead to 
# infinite results.
plot(x,y)
```
  
## 2. Sorted data Vs quantiles

```{r}
# sorting the data, storing and assigning positions
dp53_sorted <- sort(dp53$p53)
data <- data.frame(dp53$p53, dp53_sorted)
data$position <- 1:nrow(data)
n <- nrow(data)

# calculating percentiles 
data$Phi  <-  (data$position-0.5)/(n)
data$PhiInv<- qnorm(data$Phi)
# qnorm: given an area, finds the boundary value that determines this area.

# observing the data
kable(data)
plot(data$PhiInv, data$dp53_sorted)
```

## 3, 4:

```{r,fig.align ='center', fig.width = 6, fig.height = 6}
#first Manual and qqnorm plots
par(mfrow = c(1,2)) 

plot(data$PhiInv, data$dp53_sorted, 
     xlab = "Theoretical Quantiles", ylab = "Sorted p53 data", 
     main = "Manual Normal Q-Q plot")  

# Using qqnorm
qqnorm(dp53$p53, main = "qqnornm Q-Q plot" )

```

## 5

```{r}
# adding qqline
par(mfrow = c(1,3)) 
plot(data$PhiInv, data$dp53_sorted, xlab="Theoretical Quantiles",
 ylab="Sorted p53 data", main="Manual line example")

datos.cuartiles <- quantile(data$dp53_sorted, c(0.25,0.75))
norm.cuartiles <- qnorm(c(0.25, 0.75))
b <- (datos.cuartiles[2] - datos.cuartiles[1] )/(norm.cuartiles[2] - 
                                                   norm.cuartiles[1])
a <- datos.cuartiles[1] - norm.cuartiles[1] * b
abline(a, b)

qqnorm(data$dp53_sorted, main = "")
title("Qqnorm & qqline")
qqline(data$dp53_sorted)

qqPlot(data$dp53_sorted, 
       main = "qqPlot")
```

