---
title: "Exercise 4"
author: "Blanca Lizarbe"
date: "2023-10-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Observe the following data set, where $S_i$ represents the number of cells of different cellular colonies of bacteria, administered with 4 different kind of sugar each (S1, S2 or s3) 10 colonies (which means 10 groups of cells) measured for each type of sugar administration.

```{r create_p53, echo= TRUE, results='hide'}
set.seed(1)
cell.counts <- data.frame(S1 = round(rnorm(10, 16.8*10^4, 3.1*10^3)), 
                        S2 = round(rnorm(10, 13.8*10^4, 2.9*10^2)), 
                        S3 = round(rnorm(10, 19.8*10^4, 2.7*10^2)))
```
Rearrange the data frame in the most convenient way to:  

• Plot the data in a representative way.  

• Calculate a statistic to evaluate if administering S1, S2 or S3 has influence on cellular replication (cell counts). Provide the t, df and p-value for each type of comparison (independently for each) 

• Provide a CI graph of the t-tests.

• Which sugars you might say alter differently the cell counts?


